---
title: "practical_exercise_1, Methods 3, 2021, autumn semester"
author: 'Mina Almasi (Github @MinaAlmasi)'
date: "20/09/21"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 3) Brushing up on the General Linear Model

We'll do a light start and get you back in the game of thinking about formulae and how to build your linear models  
Finally, we'll have a few exercises, finishing off today's practical exercises 

## A list of formulae
```{r, eval=FALSE}
formula <- y ~ x ## y as a function of x
y ~ 1 ## model the intercept for "y"
y ~ x ## model the main effect of x and the intercept for y
y ~ x + 1 ## the same as above (+ 1 is implicit)
y ~ x + 0 ## model the main effect of x and no intercept
y ~ x - 1 ## the same as above
y ~ 0 ## doesn't model anything (for completeness)
y ~ x + z ## model the main effects x and z (and an intercept)
y ~ x:z ## model interaction of x and z
y ~ x * z ## model the main effects x and z and their interaction
y ~ x + z + x:z ## the same as above
```

## Dataset mtcars
Let's look at the "mtcars" data:  

[, 1]   mpg   Miles/(US) gallon  
[, 2]	 cyl	 Number of cylinders  
[, 3]	 disp	 Displacement (cu.in.)  
[, 4]	 hp	 Gross horsepower  
[, 5]	 drat	 Rear axle ratio  
[, 6]	 wt	 Weight (lb/1000)  
[, 7]	 qsec	 1/4 mile time  
[, 8]	 vs	 V/S  
[, 9]	 am	 Transmission (0 = automatic, 1 = manual)  
[,10]	 gear	 Number of forward gears  
[,11]	 carb	 Number of carburetors_  


## Miles per gallon and weight

We can do a scatter plot, and it looks like there is some relation between fuel usage and the weight of cars.
Let's investigate this further

```{r,fig.height=5, fig.width=6}
par(font.lab=2, font.axis=2, cex=1.2)
plot(mpg ~ wt, data=mtcars, xlab='Weight (lb/1000)', ylab='Miles/(US) gallon',
     main='Scatter plot', ylim=c(0, 40))
```

# Exercises and objectives
The objectives of today's exercises are:  
1) To remind you of the (general) linear model, and how we can use it to make models in R  
2) To make some informal model comparisons  
3) To estimate models based on binomially distributed data  

If you would like to read more about a given function, just prepend the function with a question mark, e.g.  
``` {r, eval=FALSE}
?lm
```

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below   

## Exercise 1
The general linear model: $Y = X \beta + \epsilon$:  
Do a linear regression, expressing fuel usage as a function of weight using the function __lm__  
```{r, eval=FALSE}
#Loading the data 
data(mtcars)

#Creating the linear model with fuel as a function of weight
model <- lm(formula= mpg~wt, data=mtcars)

summary(model) 
```

1. extract $\hat{\beta}$, $Y$, $\hat{Y}$, $X$ and $\epsilon$ from __model__ (hint: have a look at the function __model.matrix__)  
    i. create a plot that illustrates $Y$ and $\hat{Y}$ (if you are feeling ambitious, also include $\epsilon$ (hint: you can use the function __arrows__))

```{r exercise 1.1}
#Estimated betas
betas <- model$coefficients

betas

#Design matrix
X <- model.matrix(model)

#estimated y-values (yhat)
estimated_y <- predict(model)

#actual y-values
actual_y <- mtcars$mpg

#residuals
residuals <- model$residuals

#Plotting actual y-values vs. estimated y-values with residuals represented with the arrows function
plot(actual_y, estimated_y)
arrows(actual_y, estimated_y, x1 = actual_y, y1 = actual_y, length = 0.1, angle = 3, code = 2, col = par("fg"), lty = par("lty"), lwd = par("lwd"))
title("Actual Y-values and Estimated Y-values")

#Plotting actual y-values vs. estimated in ggplot 
intercept <- as.numeric(betas[1])
slope <- as.numeric(betas[2])

linear_plot <- ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point() +
  geom_abline(intercept=intercept, slope=slope) + 
  geom_linerange(aes(residuals), ymin = mtcars$mpg, ymax = estimated_y, x=mtcars$wt, color = "indianred", alpha = 0.5) + 
  labs(x='wt', y='mpg')

linear_plot
```

Arrows() documentation: https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/arrows

arrows(x0, y0, x1 = x0, y1 = y0 ....) 

x0, y0 -> coordinates from which to draw 

x1, y1 --> coordinates of points to which to draw. At least one must the supplied

    
2. estimate $\beta$ for a quadratic model ($y = {\beta}_{2} x^2 + {\beta}_{1} x + {\beta}_{0}$) using ordinary least squares _without_ using __lm__; $\hat{\beta} = {({X}^{T} X)}^{-1} {X}^{T} Y$ (hint: add a third column to $X$ from step 1)

```{r exercise 1.2}
#Adding a third column to X
X_NEW <- as.data.frame(X)
X_NEW$wt_squared <- X_NEW$wt^2 #adding a new column (with the x-values squared)
X_NEW <- as.matrix(X_NEW) 

#Estimating bhat for a quadratic model using the OLS ESTIMATOR
OLS_betas_q <- solve(t(X_NEW) %*% X_NEW) %*% t(X_NEW) %*% actual_y
```


3. compare your acquired $\hat{\beta}$ with the output of the corresponding quadratic model created using __lm__ (hint: use the function __I__, see details under help and the sub-section formula operators here: https://www.datacamp.com/community/tutorials/r-formula-tutorial)  
    i. create a plot that illustrates $Y$ and $\hat{Y}$ (if you are feeling ambitious, also include $\epsilon$ (hint: you can use the function __arrows__))  
    
```{r exercise 1.3}
#Creating the quadratic model using the function I()
quadratic_model <-  lm(mpg~wt+I(wt^2), data=mtcars)

betas_q <- quadratic_model$coefficients

#Comparing coefficients
all_coefficients <- cbind(OLS_betas_q, betas_q) #c-bind to combine the two vectors
colnames(all_coefficients) <- c("OLS", "lm()") #adding colnames 

all_coefficients
```

Solely by looking at the "coefficients" output, we can note that the lm() function and the ordinary least squares (OLS) estimator gives exactly the same output at 6 decimals. 


```{r exercise 1.3i}
#Estimating the y-values (yhat) using the quadratic model 
estimated_y_q <- predict(quadratic_model)

#residuals 
residuals_q <- residuals(quadratic_model)

#Plotting actual y-values vs. estimated in ggplot 
quadratic_plot <- ggplot(mtcars, aes(wt))+
  geom_point(aes(y=mpg), color = "black")+
  geom_line(aes(y=estimated_y_q), color ="black")+
  geom_linerange(aes(residuals_q), ymin = mtcars$mpg, ymax = estimated_y_q, x=mtcars$wt, color = "indianred")+
  xlab("wt") + ylab("mpg")

quadratic_plot
```


## Exercise 2
Compare the plotted quadratic fit to the linear fit  
```{r exercise 2}
#Showing the fits 
plotted_fits <- ggplot(mtcars, aes(wt))+
  geom_point(aes(y=mpg), color = "black")+
  geom_line(aes(y = estimated_y), color = "red")+
  geom_line(aes(y=estimated_y_q), color ="blue")+
  labs(x = "wt", y = "mpg", title = "Linear vs Quadratic Fit")+ 
  theme_bw()

plotted_fits

#Showing the lines separately 
ggpubr::ggarrange(linear_plot, quadratic_plot)
```

1. which seems better?  
Visually, it is difficult to tell which is better. It does not seem to be a drastic difference between the two models. 

2. calculate the sum of squared errors, (show the calculation based on $\epsilon$). Which fit has the lower sum?  

We can calculate the sum of squared errors with the following formula: 
$SSE = \sum(\hat{y_i}-y_i)^2$

where $\hat{y_i}-y_i$ represents the residuals which was calculated with the function residuals() during exercise 1. 

```{r exercise 2.2}
SSE_linear <- sum((residuals)^2)

SSE_quadratic <- sum((residuals_q)^2)

#Comparing SSE 
SSE_all <- cbind(SSE_linear, SSE_quadratic) #c-bind to combine the two vectors
colnames(SSE_all) <- c("Linear", "Quadratic") #adding colnames 

SSE_all
```

The quadratic model has the lower SSE by quite a lot. 

3. now make a cubic fit ($y = {\beta}_{3} x^3 + {\beta}_{2} x^2 + {\beta}_{1} x + {\beta}_{0}$) and compare it to the quadratic fit
    i. create a plot that illustrates $Y$ and $\hat{Y}$ for both the cubic and the quadratic fits (plot them in the same plot)  
    ii. compare the sum of squared errors  
    iii. what's the estimated value of the "cubic" (${\beta}_3$) parameter? Comment on this!  
    
```{r exercise 2.3}
#Creating the cubic model using the function I()
cubic_model <- lm(mpg~wt+I(wt^2)+I(wt^3), data=mtcars)

summary(cubic_model)
```
    
```{r exercise 2.3.i}
#Estimating the y-values (yhat) for the cubic model 
estimated_y_cubic <- predict(cubic_model)

#Plotting actual y-values vs. estimated in ggplot 
plotted_fits <- ggplot(mtcars, aes(wt))+
  geom_point(aes(y=mpg), color = "black")+
  geom_line(aes(y=estimated_y_q), color ="red")+
  geom_line(aes(y=estimated_y_cubic), color = "green")
  labs(x = "wt", y = "mpg", title = "Quadratic Fit vs. Cubic")+ 
  theme_bw()

plotted_fits
```

Visually, the two are very similar. The green line (the cubic fit) almost covers the red (the quadratic line)

```{r exercise 2.3.ii}
#cubic model residuals 
residuals_c <- residuals(cubic_model)

#cubic model SSE
SSE_cubic <- sum((residuals_c)^2)

#Comparing SSE 
SSE_qc <- cbind(SSE_quadratic, SSE_cubic) #c-bind to combine the two vectors
colnames(SSE_qc) <- c("Quadratic", "Cubic") #adding colnames 

SSE_qc
```

Numerically, the two fits are basically identical. The cubic is slightly lower, but there is barely a difference. 

```{r exercise 2.3iii}
#Printing all coefficients 
cubic_model$coefficients

#Printing beta_3 
cubic_model$coefficients[4]
```
$\hat{\beta_3}$ is very very small compared to the others estimated coefficients which may explain why the quadratic and cubic model is so similar. 

4. bonus question: which summary statistic is the fitted value (_Intercept_ or ${\beta}_0$ in $y = {\beta}_0$) below identical to?
```{r, echo=FALSE}
lm(mpg ~ 1, data=mtcars)
```
The model above is also called the *null model* which has fitted the model to a constant. 

## Exercise 3
Doing a logistic regression - estimating the probability that a car has automatic transmission (0) or manual transmission (1) based on its weight
```{r, eval=FALSE}
data(mtcars)
logistic.model <- glm(formula= am~ wt, data=mtcars, family='binomial')
```

Probabilities live on the range $(0, 1)$ - using the so-called logit function as a "link-function" we can map these onto the range $(-\infty, \infty)$, i.e. the real numbers.  
  
What we model in this case is: $Pr(y = 1) = logit^{-1}(X \beta)$, i.e. the probability of a car having manual transmission, given its weight. $X \beta$ is called the linear predictor; compare with $Y = X \beta + \epsilon$ 
It is helpful to define the logit function and its inverse function for the following:  

```{r}
logit <-     function(x) log(x / (1 - x))
inv.logit <- function(x) exp(x) / (1 + exp(x))
```

1. plot the fitted values for __logistic.model__:  
    i. what is the relation between the __linear.predictors__ and the __fitted_values__ of the __logistic.model__ object?
2. plot the logistic function, you've estimated based on your $\hat{\beta}$, (not just the fitted values). Use an _xlim_ of (0, 7)
    i. what's the interpretation of the estimated $\hat{\beta}_0$ (the _Intercept_)
    ii. calculate the estimated probability that the Pontiac Firebird has automatic transmission, given its weight
    iii. bonus question - plot the logistic function and highlight all the cars where we guessed wrongly, if we used the following "quantizer" function:
    
\begin{equation}
  transmission_{guess}=
  \begin{cases}
    1 (manual), & \text{if}\ PR(y = 1) ≥ 0.5 \\
    0 (automatic), & \text{otherwise}
  \end{cases}
\end{equation}    
    
3. plot quadratic fit alongside linear fit  
    i. judging visually, does adding a quadratic term make a difference?
    ii. check the details in the help of the AIC function - which of the models provide the better fit according to the AIC values and the residual deviance respectively?
    iii. in your own words, why might it be good to penalise a model like the quadratic model, we just fitted.
    
# Next time
We are going to looking at extending our models with so called random effects. We need to install the package "lme4" for this. Run the code below or install it from your package manager (Linux)  
```{r, eval=FALSE}
install.packages("lme4")
```
We can fit a model like this:

```{r}
library(lme4)
mixed.model <- lmer(mpg ~ wt + (1 | cyl), data=mtcars)
```

They result in plots like these:
```{r}
par(font.lab=2, font.axis=2, cex=1.2)
plot(mtcars$wt, fitted.values(mixed.model),
     main='Linear regression with group intercepts (n cylinders)',
    xlab='Weight (lb/1000)', ylab='Miles/(US) gallon',
    pch=3)
```

and this
```{r}
mixed.model <- lmer(mpg ~ wt + (wt | cyl), data=mtcars)
plot(mtcars$wt, fitted.values(mixed.model),
     main='Linear regression with group intercepts and group slopes (n cylinders)',
    xlab='Weight (lb/1000)', ylab='Miles/(US) gallon',
    pch=3)
``` 

but also new warnings like:  

Warning:
In checkConv(attr(opt, "derivs"), opt\$par, ctrl = control$checkConv,  :
  Model failed to converge with max|grad| = 0.0121962 (tol = 0.002, component 1)
